\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-12-09 misc/JSS article/R&R/article.tex
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

 Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}). The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.

@@

Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. I also wrote a blog post on \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{aggregating survey data using \pkg{collapse}}, which showcases more aspects of the \fct{collap} function using real census data. Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{A recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

pivots, and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+    fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+    fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels (stored in \code{attr(column, "label")}) are saved in an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+                   labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}. These carry out computations 'on the fly' and are thus extremely fast. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
+       names = list(from = "Variable", to = "Sectorcode"),
+       labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

This exemplifies the power of these tools to create, query, and combine nested data in very general ways, and with many applications.
  % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes.
%
\section{Summary statistics} \label{sec:summ_stat}
%
\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the indexed World Development Panel (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}. The function \fct{varying} can be used to examine which of these variables are time-varying.
%
\begin{Schunk}
\begin{Sinput}
R> varying(wlddev, ~ iso3c)
\end{Sinput}
\begin{Soutput}
      country          date          year        decade        region 
        FALSE          TRUE          TRUE          TRUE         FALSE 
       income          OECD         PCGDP        LIFEEX          GINI 
        FALSE         FALSE          TRUE          TRUE          TRUE 
          ODA           POP  center_PCGDP center_LIFEEX 
         TRUE          TRUE          TRUE          TRUE 
\end{Soutput}
\end{Schunk}
%
A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions supporting \class{indexed\_series} from Table~\ref{tab:TSfun}, this is easily demonstrated.
%
\begin{Schunk}
\begin{Sinput}
R> LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
R> all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\end{Schunk}
%
The \fct{qsu} (quick-summary) function provides an efficient method to (approximately) compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the mean back to the within transformation to preserve the scale of the data.
%
\begin{Schunk}
\begin{Sinput}
R> qsu(LIFEEXi)
\end{Sinput}
\begin{Soutput}
             N/T     Mean       SD      Min      Max
Overall    11670  64.2963  11.4764   18.907  85.4171
Between      207  64.9537   9.8936  40.9663  85.4171
Within   56.3768  64.2963   6.0842  32.9068  84.4198
\end{Soutput}
\end{Schunk}
%
The decomposition above implies more variation in life expectancy between countries than within countries over time. It can also be computed for different subgroups, such as OECD members and non-members, and with sampling weights, such as population.
%
\begin{Schunk}
\begin{Sinput}
R> qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP) |> aperm()
\end{Sinput}
\begin{Soutput}
, , FALSE

@@

%
While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very utile with mixed-type data. A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis. \newline

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}. It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
%
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE          45         30                  47                  58
  TRUE           34          0                   0                   2
\end{Soutput}
\end{Schunk}
This shows the total population (latest post-2015 estimates) in millions.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       93.01     694.89             3063.54             2459.71
  TRUE      1098.75       0.00                0.00              211.01
\end{Soutput}
\end{Schunk}
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       78.75      62.81               68.30               73.81
  TRUE        81.09                                              76.37
\end{Soutput}
\end{Schunk}
Finally, this calculates a population-weighted average of life expectancy in each group.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
+                      wFUN.args = list(w = POP)))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       77.91      63.81               68.76               75.93
  TRUE        81.13                                              76.10
\end{Soutput}
\end{Schunk}
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above functions, \pkg{collapse} also provides functions \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op}. Its contents can be accessed using \fct{get\_collapse}. \newline

@@

For deeper engagement with \pkg{collapse}, visit its \href{https://sebkrantz.github.io/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}. Users can also follow \pkg{collapse} on \href{https://x.com/collapse_R}{Twitter/X} and \href{https://bsky.app/profile/rcollapse.bsky.social}{Bluesky} to be notified about major updates and participate in community discussions.

>>>>

 Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}). The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.

@@

Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. I also wrote a blog post on \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{aggregating survey data using \pkg{collapse}}, which showcases more aspects of the \fct{collap} function using real census data. Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{A recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

pivots, and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+    fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+    fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels (stored in \code{attr(column, "label")}) are saved in an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+                   labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}. These carry out computations 'on the fly' and are thus extremely fast. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
+       names = list(from = "Variable", to = "Sectorcode"),
+       labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
More features of \fct{pivot} are demonstrated in the \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

This exemplifies the power of these tools to create, query, and combine nested data in very general ways, and with many applications.
  % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes.
%
\section{Summary statistics} \label{sec:summ_stat}
%
\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the indexed World Development Panel (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}. The function \fct{varying} can be used to examine which of these variables are time-varying.
%
\begin{Schunk}
\begin{Sinput}
R> varying(wlddev, ~ iso3c)
\end{Sinput}
\begin{Soutput}
      country          date          year        decade        region 
        FALSE          TRUE          TRUE          TRUE         FALSE 
       income          OECD         PCGDP        LIFEEX          GINI 
        FALSE         FALSE          TRUE          TRUE          TRUE 
          ODA           POP  center_PCGDP center_LIFEEX 
         TRUE          TRUE          TRUE          TRUE 
\end{Soutput}
\end{Schunk}
%
A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions supporting \class{indexed\_series} from Table~\ref{tab:TSfun}, this is easily demonstrated.
%
\begin{Schunk}
\begin{Sinput}
R> LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
R> all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\end{Schunk}
%
The \fct{qsu} (quick-summary) function provides an efficient method to (approximately) compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the mean back to the within transformation to preserve the scale of the data.
%
\begin{Schunk}
\begin{Sinput}
R> qsu(LIFEEXi)
\end{Sinput}
\begin{Soutput}
             N/T     Mean       SD      Min      Max
Overall    11670  64.2963  11.4764   18.907  85.4171
Between      207  64.9537   9.8936  40.9663  85.4171
Within   56.3768  64.2963   6.0842  32.9068  84.4198
\end{Soutput}
\end{Schunk}
%
The decomposition above implies more variation in life expectancy between countries than within countries over time. It can also be computed for different subgroups, such as OECD members and non-members, and with sampling weights, such as population.
%
\begin{Schunk}
\begin{Sinput}
R> qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP) |> aperm()
\end{Sinput}
\begin{Soutput}
, , FALSE

@@

%
While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very utile with mixed-type data. A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console. The \href{https://fastverse.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis. \newline

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}. It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
%
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE          45         30                  47                  58
  TRUE           34          0                   0                   2
\end{Soutput}
\end{Schunk}
This shows the total population (latest post-2015 estimates) in millions.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       93.01     694.89             3063.54             2459.71
  TRUE      1098.75       0.00                0.00              211.01
\end{Soutput}
\end{Schunk}
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       78.75      62.81               68.30               73.81
  TRUE        81.09                                              76.37
\end{Soutput}
\end{Schunk}
Finally, this calculates a population-weighted average of life expectancy in each group.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
+                      wFUN.args = list(w = POP)))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       77.91      63.81               68.76               75.93
  TRUE        81.13                                              76.10
\end{Soutput}
\end{Schunk}
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above functions, \pkg{collapse} also provides functions \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op}. Its contents can be accessed using \fct{get\_collapse}. \newline

@@

For deeper engagement with \pkg{collapse}, visit its \href{https://fastverse.github.io/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}. Users can also follow \pkg{collapse} on \href{https://x.com/collapse_R}{Twitter/X} and \href{https://bsky.app/profile/rcollapse.bsky.social}{Bluesky} to be notified about major updates and participate in community discussions.
